---
title: "Create Datasets for Analysis"
author: "Alex Cope"
date: '2022-11-16'
output: html_document
---

For the relevant PANSE input file, look at the date and search for the section based on the header. File paths may need to be altered. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(broom)

filterDataset <- function(input,
                          output_dir,
                          output_rfp,
                          phi,
                          output_phi,
                          min_length_gene = NULL,
                          trim_5 = NULL,
                          trim_3 = NULL,
                          final_position = NULL,
                          min_read_counts = NULL,
                          min_read_counts_per_pos = NULL,
                          apply_filter_to_front_back = F,
                          remove_outliers = FALSE,
                          exclude_these_genes = NULL,
                          num_genes = NULL,
                          pause_site_z_score_cutoff=3.92,
                          ramp.region.cutoff=0)
{
  if (!is.null(min_read_counts) && !is.null(min_read_counts_per_pos)){
    stop("only one of min_read_counts and min_read_counts_per_pos should be specified")
  }
  
  if (!is.null(trim_5) && !is.null(final_position))
  {
    print("WARNING: both final_position and trim_5 specified. final_position cutoff will be applied first, followed by additional trimming from the 5'-end.")
  }
  
  dir.create(output_dir)
  rfp <- read_csv(input)
  phi <- read_csv(phi)
  
  if (!"GeneID" %in% colnames(rfp))
  {
    rfp <- rfp %>% dplyr::rename(GeneID=gene) # some files use gene instead of GeneID
  }
  
  if (is.character(exclude_these_genes))
  {
    exclude_these_genes <- readLines(exclude_these_genes)
    rfp <- rfp %>% filter(!GeneID %in% exclude_these_genes)
  }

  
  if (!is.null(min_read_counts))
  {
    if (!apply_filter_to_front_back)
    {
      total.counts <- rfp %>% 
        group_by(GeneID) %>% 
        summarize(Total=sum(RFPCount)) %>% 
        filter(Total >= min_read_counts) %>% 
        dplyr::select(GeneID) %>% 
        deframe()
      rfp <- rfp %>% filter(GeneID %in% total.counts)
    } else {
      total.counts <- rfp %>% 
        group_by(GeneID) %>%
        mutate(Loc = ifelse(Position/n() <= 0.5,"Front","Back")) %>%
        group_by(GeneID,Loc) %>%
        summarize(Length = n(),
                  Total.RFP = sum(RFPCount),
                  RFPCount.Per.Pos = Total.RFP/Length,
        ) %>%
        pivot_wider(id_cols=c(GeneID),names_from = Loc,values_from=c(Total.RFP,Density)) %>%
        filter(Total.RFP_Front >= min_read_counts & Total.RFP_Back >= min_read_counts) %>%
        dplyr::select(GeneID) %>%
        deframe()
      rfp <- rfp %>% filter(GeneID %in% total.counts)
    }
  }
  
  if (!is.null(min_read_counts_per_pos))
  {
    if (!apply_filter_to_front_back)
    {
    total.counts <- rfp %>% 
        group_by(GeneID) %>% 
        summarize(Total=sum(RFPCount),Length=n(),Average=Total/Length) %>% 
        filter(Average >= min_read_counts_per_pos) %>% 
        dplyr::select(GeneID) %>% 
        deframe()
      rfp <- rfp %>% filter(GeneID %in% total.counts)
    } else {
      total.counts <- rfp %>% 
        group_by(GeneID) %>%
        mutate(Loc = ifelse(Position/n() <= 0.5,"Front","Back")) %>%
        group_by(GeneID,Loc) %>%
        summarize(Length = n(),
                  Total.RFP = sum(RFPCount),
                  RFPCount.Per.Pos = Total.RFP/Length,
        ) %>%
        pivot_wider(id_cols=c(GeneID),names_from = Loc,values_from=c(Total.RFP,Density)) %>%
        filter(RFPCount.Per.Pos_Front >= min_read_counts_per_pos & RFPCount.Per.Pos_Back >= min_read_counts_per_pos) %>%
        dplyr::select(GeneID) %>%
        deframe()
      rfp <- rfp %>% filter(GeneID %in% total.counts)
    }
    
  }
  
  if (remove_outliers)
  {
    rfp.split <- rfp %>% 
      group_by(GeneID) %>%
      mutate(RFPDensity = RFPCount/mean(RFPCount)) %>%
      split(f = as.factor(.$GeneID))
    
    cor.pos.rfp <- lapply(rfp.split,function(x){
        cor.test(x$Position,x$RFPDensity,method="spearman",use="complete.obs") %>% tidy()
    }) %>% bind_rows(.id="GeneID")
  
    cor.pos.rfp <- cor.pos.rfp %>%
      arrange(desc(estimate))
    
    rfp <- rfp %>% 
      group_by(GeneID) %>% 
      mutate(Is.zero = ifelse(RFPCount == 0,1,0),
             Rel.Pos = Position/Position[which.max(Position)],
             Rel.End = Position[which.max(Position)] - (Position - 1),
             Pos.Group = case_when(
                Rel.Pos <= 0.10 ~ "5_end",
                Rel.Pos >= 0.75 ~ "3_end",
                TRUE ~ "Otherwise"
                )
      )
    
    rfp.split <- rfp %>% 
      split(~as.factor(GeneID))
    
    bn.test.result <- lapply(rfp.split,function(df){
      df.sum <- df %>% 
        group_by(Pos.Group) %>% 
        summarize(Missing=sum(Is.zero),
                  Total = n())
      x <- df.sum %>% filter(Pos.Group == "5_end")
      ow <- df.sum %>% filter(Pos.Group == "3_end")
      binom.test(x$Missing,n=x$Total,p=ow$Missing/ow$Total,alternative="greater") %>% tidy()
    }) %>% bind_rows(.id="GeneID")
  
    
    rfp <- rfp %>%
      group_by(GeneID) %>%
      mutate(RFPDensity = RFPCount/mean(RFPCount,na.rm=T),
             Log.RFPDensity = log(RFPDensity),
             Log.RFPDensity = ifelse(is.finite(Log.RFPDensity),Log.RFPDensity,NA),
             Z = (Log.RFPDensity - mean(Log.RFPDensity,na.rm = T))/sd(Log.RFPDensity,na.rm=T)
      )
    
    
    z.score.cutoff <- rfp %>% 
      filter(Z > pause_site_z_score_cutoff) %>% 
      dplyr::select(GeneID) %>% 
      unique()  %>% 
      deframe()
    
    cor.cutoff <- quantile(cor.pos.rfp$estimate,probs=c(0.05,0.95))
    cor.pos.rfp.cutoff <- cor.pos.rfp %>%
      filter(estimate <= cor.cutoff[1] | estimate >= cor.cutoff[2]) %>% 
      dplyr::select(GeneID) %>% 
      unique() %>% 
      deframe()
    
    binom.cutoff <- bn.test.result %>% 
      filter(p.value < 0.05) %>% 
      dplyr::select(GeneID) %>% 
      unique()  %>% 
      deframe()
    
    to.remove <-unique(c(z.score.cutoff,cor.pos.rfp.cutoff,binom.cutoff))
    rfp <- rfp %>%
      filter(!GeneID %in% to.remove) %>%
      dplyr::select(GeneID,Position,Codon,RFPCount)
  }
  
  if (!is.null(final_position))
  {
    rfp <- rfp %>% 
      filter(Position <= final_position)
  }
  
  if (!is.null(trim_5))
  {
    rfp <- rfp %>% 
      filter(Position > trim_5) %>% 
      mutate(Position = Position - trim_5)
  }
  
  if (!is.null(trim_3))
  {
    length.per.gene <- rfp %>% 
      group_by(GeneID) %>%
      summarize(Length=n())
    rfp <- rfp %>% 
      left_join(length.per.gene,by="GeneID")
    rfp <- rfp %>% 
      mutate(Rel.to.3.end = Length - Position) %>%
      filter(Rel.to.3.end > trim_3) %>%
      dplyr::select(-Length,-Rel.to.3.end)
  }
  
  if(!is.null(num_genes))
  {
    gene.names <- unique(rfp$GeneID)
    num.genes.remaining <- length(gene.names)
    if (num.genes.remaining > num_genes)
    {
      genes.to.keep <- sample(gene.names,size=num_genes,replace = F)
      rfp <- rfp %>% 
        filter(GeneID %in% genes.to.keep)
    }
  }
  if (!is.null(min_length_gene))
  {
    short <- rfp %>% 
      group_by(GeneID) %>% 
      dplyr::count() %>% 
      filter(n < min_length_gene)
    
    rfp <- rfp %>% 
      filter(!GeneID %in% short$GeneID)
  }
  
  rfp <- rfp %>%
    mutate(Mixture = ifelse(Position <= ramp.region.cutoff,-1,1)) %>%
    dplyr::select(GeneID,Position,Codon,Mixture,RFPCount)
  
  if ("ORF" %in% colnames(phi))
  {
    phi <- phi %>% column_to_rownames("ORF")
  } else{
    phi <- phi %>% column_to_rownames("GeneID")
  }
  phi <- phi[unique(rfp$GeneID),,drop=F]
  phi <- phi %>% rownames_to_column("GeneID")
  
  print(str_c("Remaining number of genes: ",nrow(phi)))
  write_csv(rfp,file=file.path(output_dir,output_rfp))
  write_csv(phi,file=file.path(output_dir,output_phi))
}

```

## Weinberg

To test the effects of key filtering parameters, turn all filtering parameters off other than filtering bad sequences.

## 04-21-2023

```{r}
input <- "/data2/cope/Nonsense_error_rates/Data/PANSE_Formatted/Scerevisiae/Weinberg_etal_2016/weinberg_etal_2016_all_frames.csv"
output_dir <- "/data2/cope/Nonsense_error_rates/Data/PANSE_Formatted/Scerevisiae/Weinberg_etal_2016/All_frames/Full_genome/"
output_rfp <- "2023-04-21_weinberg_etal_2016_all_frames.csv"
phi <- "/data2/cope/Nonsense_error_rates/Data/PANSE_Formatted/Scerevisiae/Weinberg_etal_2016/weinberg_etal_2016_all_frames_phi.csv"
output_phi <- "2023-04-21_weinberg_etal_2016_all_frames_phi.csv"
min_length_gene <- 0
min_read_counts <- 1
remove_outliers <- TRUE
exclude_these_genes <- "/data2/cope/Nonsense_error_rates/Data/proteins_to_exclude_paralogs_mito_scer.txt"
#exclude_these_genes <- NULL

filterDataset(input = input,
                      output_dir = output_dir,
                      output_rfp = output_rfp,
                      phi = phi,
                      output_phi = output_phi,
                      min_length_gene = min_length_gene,
                      min_read_counts = min_read_counts,
                      remove_outliers = remove_outliers,
                      exclude_these_genes = exclude_these_genes)

data <- read_csv("/data2/cope/Nonsense_error_rates/Data/PANSE_Formatted/Scerevisiae/Weinberg_etal_2016/All_frames/Full_genome/2023-04-21_weinberg_etal_2016_all_frames.csv")
for (i in seq(0,200))
{
  data.w.mix <- data %>% 
    mutate(Mixture = ifelse(Position < i,-1,1)) %>%
    dplyr::select(GeneID,Position,Codon,Mixture,RFPCount)
  filename <- paste0("2023-04-21_weinberg_etal_2016_all_frames_",i,"_ramp.csv")
  write_csv(data.w.mix,file=file.path(output_dir,filename))
}


```




## 06-12-2023

```{r}
rfp <- read_csv("../Data/PANSE_Formatted/Scerevisiae/Weinberg_etal_2016/All_frames/Full_genome/2023-05-08_weinberg_etal_2016_all_frames_0_ramp.csv")
phi <- read_csv("../Data/PANSE_Formatted/Scerevisiae/Weinberg_etal_2016/All_frames/Full_genome/2023-05-08_weinberg_etal_2016_all_frames_phi.csv")
rfp.filt <- rfp %>%
  group_by(GeneID) %>%
  filter(Position[which.max(Position)] >= 225)
phi.filt <- phi %>%
  filter(GeneID %in% unique(rfp.filt$GeneID))

rfp.filt.ramp <- rfp.filt %>%
  mutate(Mixture = ifelse(Position <= 200,-1,1))

write_csv(rfp.filt,"../Data/PANSE_Formatted/Scerevisiae/Weinberg_etal_2016/All_frames/Full_genome/2023-06-12_weinberg_etal_2016_all_frames_0_ramp.csv")
write_csv(rfp.filt.ramp,"../Data/PANSE_Formatted/Scerevisiae/Weinberg_etal_2016/All_frames/Full_genome/2023-06-12_weinberg_etal_2016_all_frames_200_ramp.csv")
write_csv(phi.filt,"../Data/PANSE_Formatted/Scerevisiae/Weinberg_etal_2016/All_frames/Full_genome/2023-06-12_weinberg_etal_2016_all_frames_phi.csv")



```

```{r}
rfp <- read_csv("../Data/PANSE_Formatted/Scerevisiae/Weinberg_etal_2016/All_frames/Full_genome/2023-05-08_weinberg_etal_2016_all_frames_0_ramp.csv")
phi <- read_csv("../Data/PANSE_Formatted/Scerevisiae/Weinberg_etal_2016/All_frames/Full_genome/2023-05-08_weinberg_etal_2016_all_frames_phi.csv")
rfp.filt <- rfp %>%
  group_by(GeneID) %>%
  filter(Position[which.max(Position)] >= 225)
phi.filt <- phi %>%
  filter(GeneID %in% unique(rfp.filt$GeneID))

rfp.filt.ramp <- rfp.filt %>%
  mutate(Mixture = ifelse(Position <= 200,1,-1))

write_csv(rfp.filt.ramp,"../Data/PANSE_Formatted/Scerevisiae/Weinberg_etal_2016/All_frames/Full_genome/2023-07-06_weinberg_etal_2016_all_frames_ignore_post_200.csv")
write_csv(phi.filt,"../Data/PANSE_Formatted/Scerevisiae/Weinberg_etal_2016/All_frames/Full_genome/2023-07-06_weinberg_etal_2016_all_frames_phi.csv")



```




## 09-22-2025

```{r warning=FALSE}
input <- "00_unfiltered_genes/wu_etal_2019_chx1_all_genes.csv"
output_dir <- "./"
output_rfp <- "2025-09-22_wu_etal_2019_chx1_all_frames_200_ramp.csv"
phi <- "00_unfiltered_genes/wu_etal_2019_chx1_phi.csv"
output_phi <- "2025-09-22_wu_etal_2019_chx1_all_frames_phi.csv"
min_length_gene <- 225
min_read_counts <- 1
remove_outliers <- TRUE
exclude_these_genes <- "../proteins_to_exclude_paralogs_mito_scer.txt"

filterDataset(input = input,
                      output_dir = output_dir,
                      output_rfp = output_rfp,
                      phi = phi,
                      output_phi = output_phi,
                      min_length_gene = min_length_gene,
                      min_read_counts = min_read_counts,
                      remove_outliers = remove_outliers,
                      ramp.region.cutoff = 200,
                      exclude_these_genes = exclude_these_genes)


```


## 09-30-2025

```{r warning=FALSE}
input <- "00_unfiltered_genes/tomuro_etal_2024_ribopool_filter_all_genes.csv"
output_dir <- "./"
output_rfp <- "2025-09-30_tomuro_etal_2024_ribopool_filter_all_frames_200_ramp.csv"
phi <- "00_unfiltered_genes/tomuro_etal_2024_ribopool_filter_phi.csv"
output_phi <- "2025-09-30_tomuro_etal_2024_ribopoo_filter_all_frames_phi.csv"
min_length_gene <- 225
min_read_counts <- 1
remove_outliers <- TRUE
exclude_these_genes <- "../proteins_to_exclude_paralogs_mito_scer.txt"

filterDataset(input = input,
                      output_dir = output_dir,
                      output_rfp = output_rfp,
                      phi = phi,
                      output_phi = output_phi,
                      min_length_gene = min_length_gene,
                      min_read_counts = min_read_counts,
                      remove_outliers = remove_outliers,
                      ramp.region.cutoff = 200,
                      exclude_these_genes = exclude_these_genes)


```

## 10-07-2025

Going to repeat analysis of Weinberg et al

```{r warning=FALSE}
input <- "00_unfiltered_genes/weinberg_etal_2016_replicate_all_genes.csv"
output_dir <- "./"
output_rfp <- "2025-10-07_weinberg_etal_2016_all_frames_200_ramp.csv"
phi <- "00_unfiltered_genes/weinberg_etal_2016_phi.csv"
output_phi <- "2025-10-07_weinberg_etal_2016_all_frames_phi.csv"
min_length_gene <- 225
min_read_counts <- 1
remove_outliers <- TRUE
exclude_these_genes <- "../proteins_to_exclude_paralogs_mito_scer.txt"

filterDataset(input = input,
                      output_dir = output_dir,
                      output_rfp = output_rfp,
                      phi = phi,
                      output_phi = output_phi,
                      min_length_gene = min_length_gene,
                      min_read_counts = min_read_counts,
                      remove_outliers = remove_outliers,
                      ramp.region.cutoff = 200,
                      exclude_these_genes = exclude_these_genes)


```
## 10-08-2025

Going to repeat analysis of Weinberg et al with 28 length reads using 15nt offset, while 29 and 30 use a 16nt offset as determined using the riboWaltz tool

```{r warning=FALSE}
input <- "00_unfiltered_genes/weinberg_etal_2016_replicate_different_offsets_all_genes.csv"
output_dir <- "./"
output_rfp <- "2025-10-08_weinberg_etal_2016_all_frames_200_ramp.csv"
phi <- "00_unfiltered_genes/weinberg_etal_2016_phi.csv"
output_phi <- "2025-10-08_weinberg_etal_2016_all_frames_phi.csv"
min_length_gene <- 225
min_read_counts <- 1
remove_outliers <- TRUE
exclude_these_genes <- "../proteins_to_exclude_paralogs_mito_scer.txt"

filterDataset(input = input,
                      output_dir = output_dir,
                      output_rfp = output_rfp,
                      phi = phi,
                      output_phi = output_phi,
                      min_length_gene = min_length_gene,
                      min_read_counts = min_read_counts,
                      remove_outliers = remove_outliers,
                      ramp.region.cutoff = 200,
                      exclude_these_genes = exclude_these_genes)


```


## 11-04-2025

### Weinberg data
Going to repeat analysis of Weinberg et al. with putative No-Go Decay targets and genes prone to frameshifts removed 

```{r warning=FALSE}
input <- "00_unfiltered_genes/weinberg_etal_2016_replicate_different_offsets_all_genes.csv"
output_dir <- "./"
output_rfp <- "2025-11-04_weinberg_etal_2016_all_frames_200_ramp.csv"
phi <- "00_unfiltered_genes/weinberg_etal_2016_phi.csv"
output_phi <- "2025-11-04_weinberg_etal_2016_all_frames_phi.csv"
min_length_gene <- 225
min_read_counts <- 1
remove_outliers <- TRUE
exclude_these_genes <- "../proteins_to_exclude_paralogs_mito_scer_frameshift_ngd.txt"

filterDataset(input = input,
                      output_dir = output_dir,
                      output_rfp = output_rfp,
                      phi = phi,
                      output_phi = output_phi,
                      min_length_gene = min_length_gene,
                      min_read_counts = min_read_counts,
                      remove_outliers = remove_outliers,
                      ramp.region.cutoff = 200,
                      exclude_these_genes = exclude_these_genes)


```



### Ferguson

```{r warning=FALSE}
input <- "00_unfiltered_genes/ferguson_etal_2023_ottr_p1_all_genes.csv"
output_dir <- "./"
output_rfp <- "2025-11-04_ferguson_etal_2023_all_frames_200_ramp.csv"
phi <- "00_unfiltered_genes/ferguson_etal_2023_ottr_p1_all_genes_phi.csv"
output_phi <- "2025-11-04_ferguson_etal_2023_all_frames_phi.csv"
min_length_gene <- 225
min_read_counts <- 1
remove_outliers <- TRUE
exclude_these_genes <- "../proteins_to_exclude_paralogs_mito_scer.txt"

filterDataset(input = input,
                      output_dir = output_dir,
                      output_rfp = output_rfp,
                      phi = phi,
                      output_phi = output_phi,
                      min_length_gene = min_length_gene,
                      min_read_counts = min_read_counts,
                      remove_outliers = remove_outliers,
                      ramp.region.cutoff = 200,
                      exclude_these_genes = exclude_these_genes)


```


## 11-05-2025

### Wu et al. with 28nt and 21nt
Note the Wu et al. data here used A-site offsets predicted from riboWaltz.

```{r warning=FALSE}
input <- "00_unfiltered_genes/wu_etal_2019_chx1_all_genes_21nt_and_28nt_reads.csv"
output_dir <- "./"
output_rfp <- "2025-11-05_wu_etal_2019_all_frames_21nt_28nt_200_ramp.csv"
phi <- "00_unfiltered_genes/wu_etal_2019_chx1_phi.csv"
output_phi <- "2025-11-05_wu_etal_2019_all_frames_21nt_28nt_phi.csv"
min_length_gene <- 225
min_read_counts <- 1
remove_outliers <- TRUE
exclude_these_genes <- "../proteins_to_exclude_paralogs_mito_scer.txt"

filterDataset(input = input,
                      output_dir = output_dir,
                      output_rfp = output_rfp,
                      phi = phi,
                      output_phi = output_phi,
                      min_length_gene = min_length_gene,
                      min_read_counts = min_read_counts,
                      remove_outliers = remove_outliers,
                      ramp.region.cutoff = 200,
                      exclude_these_genes = exclude_these_genes)


```
## 11-07-2025
Ferguson et al. data OTTR P1 exhibits a smaller ramp region near the 5'-end (about 50 nt).

```{r warning=FALSE}
input <- "00_unfiltered_genes/ferguson_etal_2023_ottr_p1_all_genes.csv"
output_dir <- "./"
output_rfp <- "2025-11-07_ferguson_etal_2023_all_frames_50_ramp.csv"
phi <- "00_unfiltered_genes/ferguson_etal_2023_ottr_p1_all_genes_phi.csv"
output_phi <- "2025-11-07_ferguson_etal_2023_all_frames_phi.csv"
min_length_gene <- 225
min_read_counts <- 1
remove_outliers <- TRUE
exclude_these_genes <- "../proteins_to_exclude_paralogs_mito_scer.txt"

filterDataset(input = input,
                      output_dir = output_dir,
                      output_rfp = output_rfp,
                      phi = phi,
                      output_phi = output_phi,
                      min_length_gene = min_length_gene,
                      min_read_counts = min_read_counts,
                      remove_outliers = remove_outliers,
                      ramp.region.cutoff = 50,
                      exclude_these_genes = exclude_these_genes)


```

## 11-21-2025
Ferguson et al. data OTTR P1 is a bit sparser than some of our other datasets. They did two technical replicates for these measurements, so let's simply combine them by adding together the read counts assigned to each codon. 

```{r warning=FALSE}
input <- "00_unfiltered_genes/ferguson_etal_2023_ottr_p1_rep1_plus_rep2_all_genes.csv"
output_dir <- "./"
output_rfp <- "2025-11-21_ferguson_etal_2023_all_frames_50_ramp.csv"
phi <- "00_unfiltered_genes/ferguson_etal_2023_ottr_p1_rep1_plus_rep2_all_genes_phi.csv"
output_phi <- "2025-11-21_ferguson_etal_2023_all_frames_phi.csv"
min_length_gene <- 150
min_read_counts <- 1
remove_outliers <- TRUE
exclude_these_genes <- "../proteins_to_exclude_paralogs_mito_scer.txt"

filterDataset(input = input,
                      output_dir = output_dir,
                      output_rfp = output_rfp,
                      phi = phi,
                      output_phi = output_phi,
                      min_length_gene = min_length_gene,
                      min_read_counts = min_read_counts,
                      remove_outliers = remove_outliers,
                      ramp.region.cutoff = 50,
                      exclude_these_genes = exclude_these_genes,
                      pause_site_z_score_cutoff=3.92)


```