---
title: "Model Adequacy"
author: "Alexander Cope"
date: "2024-05-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(AnaCoDa)

plotting.target <- switch(1,
                          "paper",
                          "talk")
source("ggplot2.settings.R")
suppressMessages(source("helperFunctions.R"))

```


# Comparing real ribosome densities to ribosome densities simulated using parameters estimated from the PANSE model

For this analysis, we will base in on the set of filtered genes as outlined in ``00_data_quality_and_filtering.Rmd`. For simulating ribosome counts at each count, we will use the posterior means from the model fit `../01_results/00_panse_fits/2023-07-20_Weinberg_etal_all_genes_filter_genes_shorter_than_225_codons_200_ramp_run_2`.

```{r}

rfp <- read_csv("../00_data/00_panse_input/2023-06-12_weinberg_etal_2016_all_frames_200_ramp.csv")
if ("gene" %in% colnames(rfp))
{
  rfp <- rfp %>%
    rename(GeneID = gene)
}

rfp <- rfp %>%
  mutate(RFPCount.w.NA = ifelse(Position <= 200,NA,RFPCount)) %>%
  group_by(GeneID) %>%
  mutate(RFPDensity = RFPCount/mean(RFPCount.w.NA,na.rm=T),
         RFPDensity = ifelse(is.infinite(RFPDensity),NA,RFPDensity))

rfp.by.pos <- rfp %>%
  ungroup() %>%
  group_by(Position) %>%
  dplyr::summarize(Density= mean(RFPDensity,na.rm=T))

sim.rfp <- read_csv("../01_results/01_simulations/simulated_rfp.csv")

sim.rfp <- sim.rfp %>%
  mutate(RFPCount.w.NA = ifelse(Position <= 200,NA,RFPCount)) %>%
  group_by(GeneID) %>%
  mutate(RFPDensity = RFPCount/mean(RFPCount.w.NA,na.rm=T),
         RFPDensity = ifelse(is.infinite(RFPDensity),NA,RFPDensity))



sim.rfp.by.pos <- sim.rfp %>%
  ungroup() %>%
  group_by(Position) %>%
  dplyr::summarize(Density= mean(RFPDensity,na.rm=T))

metagene.comp <- rfp.by.pos %>%
  left_join(sim.rfp.by.pos,by=c("Position"),suffix=c("_Real","_Simulated")) %>%
  filter(Position <= 1000) %>%
  pivot_longer(c(Density_Real,Density_Simulated),names_to = "Data",values_to="Density") %>%
  mutate(Data = ifelse(Data== "Density_Real","Real","Simulated"))


metagene.comp.post.200 <- metagene.comp %>%
  mutate(Density = ifelse(Position <= 200,NA,Density))
metagene.comp.pre.200 <- metagene.comp %>%
  mutate(Density = ifelse(Position > 500,NA,Density))

full.ramp.plot <- ggplot(metagene.comp,aes(x=Position,y=Density,color=Data)) +
  geom_line() +
  ylab("Ribosome Density") +
  scale_color_viridis_d() +
  geom_hline(yintercept = 1,linetype="dashed") +
  xlab("Position") +
  annotate("rect",fill="black",alpha=0.4,xmin=0,xmax=200,ymin=-Inf,ymax=Inf) +
  annotate("text",x=600,y=2.5,label="Region used for\nparameter\nestimation",size=3)+
  annotate("text",x=100,y=2.5,label="Excluded",color="white",size=3)+
  theme_cowplot()
full.ramp.plot


post.200.ramp.plot <- ggplot(metagene.comp.post.200,aes(x=Position,y=Density,color=Data)) +
  geom_line() +
  ylab("Ribosome Density") +
  scale_color_viridis_d() +
  geom_hline(yintercept = 1,linetype="dashed") +
  xlab("Position") +
  ylim(NA,max(metagene.comp$Density)) +
  scale_x_continuous(breaks=c(0,200,400,600,800,1000)) +
  theme_cowplot()
post.200.ramp.plot

pre.200.ramp.plot <- ggplot(metagene.comp.pre.200,aes(x=Position,y=Density,color=Data)) +
  geom_line() +
  ylab("Ribosome Density") +
  scale_color_viridis_d() +
  geom_hline(yintercept = 1,linetype="dashed") +
  xlab("Position") +
  scale_x_continuous(breaks=c(0,500),limits=c(0,500)) +
  ylim(min(metagene.comp$Density),NA) +
  theme_cowplot() +
  geom_vline(xintercept = c(50,100,200))
pre.200.ramp.plot


```


```{r}
rfp.real.vs.sim <- rfp %>%
  left_join(sim.rfp,by=c("GeneID","Position","Codon"),suffix=c("_Real","_Sim"))

cor.test(rfp.real.vs.sim$RFPCount_Real,rfp.real.vs.sim$RFPCount_Sim,method="spearman")

rfp.real.vs.sim.no.zero <- rfp.real.vs.sim %>%
  filter(RFPCount_Real != 0 & RFPCount_Sim != 0 & Position > 200) %>%
  mutate(Deviation = log(RFPDensity_Real) - log(RFPDensity_Sim)) %>%
  filter(!is.nan(Deviation))
  

dev.hist <- ggplot(rfp.real.vs.sim.no.zero,aes(x=Deviation)) +
  geom_histogram(color="black",fill="white") +
  theme_cowplot() +
  xlab("Deviation\nLog(True RFP Count) - Log(Simulated RFP Count)") +
  ylab("Count")
dev.hist

ggsave2(file.path(fig.dir,"deviation_real_vs_simulated.pdf"),dev.hist,height = 5,width=7)

t.test(rfp.real.vs.sim.no.zero$Deviation)
print(paste("The fold change on the natural scale is:", exp(mean(rfp.real.vs.sim.no.zero$Deviation,na.rm=T))))
```

## First 200 vs. post-200

This analysis suggests that the first 200 codons appears to have highly-inflated parameter estimates relative to the the remainder of the genes.
This is consistent with previous results that applied Sin et al.'s approach to the Weinberg data.

```{r}
ignore.post.200 <- getParameterDataFrames("../01_results/00_panse_fits/2023-07-20_Weinberg_etal_all_genes_filter_genes_shorter_than_225_codons_ignore_post_200_ramp/final_restart/")
ignore.first.200 <- getParameterDataFrames("../01_results/00_panse_fits/2026-01-07_Weinberg_etal_2016_all_frames_filter_genes_shorter_than_225_codons_200_ramp_updated_posterior_variable_nse/final_restart/")

ramp.vs.post.nse <- compareEstimates(ignore.first.200,ignore.post.200,variable="Log.NSERate",ylab="First 200 Codons\nLog10(NSE Rate)",xlab="Post-200 codons\nLog10(NSE Rate)",title="",color=NA) 
  

ramp.vs.post.wt <- compareEstimates(ignore.post.200,ignore.first.200,variable="Wait.Time",xlab="Elongation Waiting Times\nFirst 200 Codons",ylab="Elongation Waiting Times\nPost-200 codons",title="",color =  NA)

ramp.vs.post.phi <- compareEstimates(ignore.post.200,ignore.first.200,variable="Phi",xlab="Total Initiation Rate\nFirst 200 Codons",ylab="Total Initiation Rate\nPost-200 codons",title="",log.scale.phi = T)

ramp.vs.post.nse
ramp.vs.post.wt
ramp.vs.post.phi


```

### Probability of NSE occurring in first 200 codons

```{r}
p <- mean(ignore.post.200$NSEProb$Mean)
k <- 1:200
sum(dgeom(k,p))


```


```{r fig.width=10,fig.height=5}
ramp.nse.effect <- (full.ramp.plot | ramp.vs.post.nse) + plot_annotation(tag_levels = "A") + plot_layout(widths = c(1.5,1))
ramp.nse.effect
ggsave2(file.path(fig.dir, "Figure_3.pdf"),ramp.nse.effect,width=11,height=4.5)

ramp.other.effect <- ramp.vs.post.wt | ramp.vs.post.phi + plot_annotation(tag_levels = "A")

ggsave2(file.path(fig.dir,"ramp_effect_waiting_time_and_phi.pdf"),ramp.other.effect,width=9,height=5)


```

